---
title: "2016 Election Polls Analysis"
author: "Ilse Dippenaar and Sam Gilbert"
output:
  html_document: 
    toc: true
    toc_depth: 2
    number_sections: true
    theme: journal
---

Poll data received from:
https://www.kaggle.com/fivethirtyeight/2016-election-polls

Election results compiled from:
http://www.cnn.com/election/results/president

The dataset we decided to analyze was polling data from the 2016 U.S. presidential election. Originally the dataset included 2 duplicates of the initial polling data, but modified to fit alternative models. In our analysis, we have filtered these additional models out. We then filtered additional unnecessary columns. The remaining variables in our dataset are as follows:

  state: The U.S. state for which the poll was predicting a winner. U.S. means it was a national poll.
  startdate: The time the poll began.
  enddate: The time the poll ended.
  pollster: The organization running the poll.
  grade: The letter grade of the quality of the poll.
  samplesize: The number of people polled.
  population: The type of people polled. In this case, "lv" = "likely voters", "rv" = "registered voters", "a" = "adults" and "v" = voters. 
  rawpoll_clinton: The raw percentage awarded to Clinton.
  rawpoll_trump: The raw percentage awarded to Trump.
  rawpoll_johnson: The raw percentage awarded to Johnson.
  poll_id: The unique integer assigned to each poll.
  
  
  For more information on the adjustments for adjpoll variables, see:
  https://fivethirtyeight.com/features/a-users-guide-to-fivethirtyeights-2016-general-election-forecast/

```{r, message=F}
library(mosaic)
library(ggplot2)
library(dplyr)
library(MASS)
library(DMwR)
library(randomForest)
library(party)
library(class)
```


# Cleaning Data
```{r}
predicted <- function(clinton, trump, johnson) {
  pred <- apply(cbind(clinton, trump, johnson), 1, which.max)
  ifelse(pred == 1, "Clinton", ifelse(pred == 2, "Trump", "Johnson"))
}
```


```{r}
make_state_results_data_frame <- function() {
  # national_2016.csv from https://econsnapshot.com/2016/11/30/our-election-data/
  county_raw <- read.csv("national_2016.csv", stringsAsFactors = F)
  state_results_raw1 <- county_raw %>%
    group_by(state) %>%
    mutate(total_votes = sum(votes)) %>%
    group_by(candidate, add = T) %>%
    # total votes will be the same across candidates
    summarize(percent = (sum(votes) / total_votes[1] * 100))
  
  state_abbrevs <- read.csv("state_abbrevs.csv", stringsAsFactors = F)
  # Initialize empty data frame with column names
  all_candidate_names <- unique(state_results_raw1$candidate)
  make_new_row <- function() {
    new_df <- data.frame(state="")
    for (candidate in all_candidate_names) {
      new_df[candidate] <- 0
    }
    return(new_df)
  }
  state_results_raw2 <- make_new_row()
  state_results_raw2 <- state_results_raw2[-1,]
  for (state_ab in unique(state_results_raw1$state)) {
    current_result <- subset(state_results_raw1, state == state_ab)
    state_name <- subset(state_abbrevs, state_abbrev == state_ab)[["state_name"]]
    next_row <- make_new_row()
    next_row["state"] <- state_name
    for (i in 1:nrow(current_result)) {
      next_row[current_result[["candidate"]][i]] <- current_result[["percent"]][i]
    }
    state_results_raw2 <- rbind(state_results_raw2, next_row)
  } 
  clinton <- apply(state_results_raw2[c("Clinton", "H. Clinton", "Hillary Clinton")], 1, sum) 
  trump <- apply(state_results_raw2[c("Trump", "D. Trump", "Donald Trump")], 1, sum) 
  johnson <- apply(state_results_raw2[c("G. Johnson", "Gary Johnson")], 1, sum) 
  other <- apply(state_results_raw2[-1], 1, sum) - clinton - trump - johnson  
  
  total_votes <- sum(county_raw$votes) 
  votes_clinton <- sum(subset(county_raw, candidate %in% c("Clinton", "H. Clinton", "Hillary Clinton"))$votes) 
  votes_trump <- sum(subset(county_raw, candidate %in% c("Trump", "D. Trump", "Donald Trump"))$votes)
  votes_johnson <- sum(subset(county_raw,
                          candidate %in% c("G. Johnson", "Gary Johnson"))$votes)
  votes_other <- total_votes - (votes_clinton + votes_trump + votes_johnson)
  
  ret <- data.frame(state=state_results_raw2[["state"]],
                    actual_clinton = clinton, 
                    actual_trump = trump, 
                    actual_johnson = johnson, 
                    actual_other = other,
                    stringsAsFactors = F)
  rbind(ret, data.frame(state = "U.S.", 
                        actual_clinton = votes_clinton / total_votes * 100,
                        actual_trump = votes_trump / total_votes * 100,
                        actual_johnson = votes_johnson / total_votes * 100,
                        actual_other = votes_other / total_votes * 100))
}
state_results <- make_state_results_data_frame()
```


```{r, message=F, warning=F}
# TODO: combine congressional districts
outcomes <- read.csv("outcomes.csv", sep=",", strip.white = T, stringsAsFactors = F)
raw <- read.csv("presidential_polls.csv", sep=",", strip.white = T, stringsAsFactors = F)
poll_quality <- data.frame(grade = c("A+", "A", "A-",
                                    "B+", "B", "B-",
                                    "C+", "C", "C-",
                                    "D+", "D", "D-",
                                    ""), 
                          grade_value = c(12:1, NA), 
                          stringsAsFactors = F)

d <- raw %>% 
  filter(type == "polls-only",
         !is.na(samplesize)) %>%
  dplyr::select(-c(cycle, branch, type, matchup, forecastdate, multiversions, timestamp, 
            rawpoll_mcmullin, adjpoll_mcmullin, question_id, createddate, url, 
            adjpoll_clinton, adjpoll_trump, adjpoll_johnson, poll_wt, population)) %>%
  left_join(outcomes, by = "state") %>%
  left_join(poll_quality, by = "grade") %>%
  left_join(state_results, by = "state") %>%
  filter(!is.na(result)) %>% # this excludes rows with state == "Nebraska CD-1" and similar rows
  mutate(predicted = predicted(rawpoll_clinton, rawpoll_trump, rawpoll_johnson),
         is_correct = predicted == result,
         duration = as.numeric(as.Date(enddate, "%m/%d/%Y") - as.Date(startdate, "%m/%d/%Y")),
         is_national = state == "U.S.",
         end_date_to_election = as.numeric(as.Date("11/09/2016", "%m/%d/%Y") - 
                                              as.Date(enddate, "%m/%d/%Y")))
d["error"] <- apply(
  d[c("rawpoll_clinton", "rawpoll_trump", "rawpoll_johnson")] - 
    d[c("actual_clinton", "actual_trump", "actual_johnson")], 1, function(row) mean(abs(row), na.rm = T))
```


# Data Exploration

```{r}
pres_colors <- c("#14B3F5", "#DD342F")
ggplot(d, aes(rawpoll_clinton, rawpoll_trump, color=result)) +
  geom_point(size = 0.5) +
  xlim(0, 100) + ylim(0, 100) +
  scale_color_manual(values = pres_colors, name = "Result") +
  labs(x = "Clinton Percent", y = "Trump Percent") +
  theme_light() +
  theme(aspect.ratio = 1)
```


```{r}
mod <- prcomp(d %>% 
                filter(!is.na(rawpoll_johnson)) %>% 
                dplyr::select(rawpoll_clinton, rawpoll_trump, rawpoll_johnson), scale. = T)
components <- mod$x
```


```{r}
ggplot(d %>% filter(!is.na(rawpoll_johnson)), aes(components[,1], components[,2], color=result)) + 
  geom_point(alpha = 1, size = 0.5) +
  scale_color_manual(values = pres_colors) +
  labs(x = "Principal Component 1", y = "Principal Component 2") +
  theme_light() +
  theme(aspect.ratio = 1)
```

```{r}
d %>%
  filter(!is.na(grade_value)) %>%
  ggplot(aes(is_correct, grade_value)) +
  geom_boxplot()
```

So, grade value has no bearing on whether or not the poll predicted the election correctly... does anything else? Maybe duration, time it started, time it ended, or location?

```{r}
summary(lm(error ~ grade_value, data = d))
```
So the error rate is not dependent on grade, but maybe incredibly slightly. 
 
 
```{r}
states <- map_data("state") %>% 
  left_join(d %>%
              filter(state != "U.S.") %>% 
              mutate(state = tolower(state),
                     poll_clinton = samplesize * rawpoll_clinton,
                     poll_trump = samplesize * rawpoll_trump) %>%
              group_by(state) %>% 
              mutate(total_samplesize = sum(samplesize)) %>%
              summarize(predicted_clinton = sum(poll_clinton) / total_samplesize[1],
                        predicted_trump = sum(poll_trump) / total_samplesize[1],
                        error = (abs(predicted_clinton - actual_clinton[1]) + 
                                 abs(predicted_trump - actual_trump[1])) / 2), # take the 1-norm of the error
            by=c("region" = "state"))
ggplot() +
  geom_polygon(data = states, aes(long, lat, group = group, fill=error), color="black") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "white", high = "red", name = "Error") +
  theme_map()
```


```{r}
table(d$predicted, d$is_correct)/nrow(d)
USpolls <- d %>% filter(state == "U.S.")
table(USpolls$predicted, USpolls$is_correct) / nrow(USpolls)
state_polls <- d %>% filter(state != "U.S.")
table(state_polls$predicted, state_polls$is_correct) / nrow(state_polls)
```


# Models

## Cross Validation function
```{r}
run_cv <- function(data, formula, 
                   response = NA, fit_fun = lm, predict_fun = predict, folds = 10, exclude_na = T) {
  if (exclude_na) data <- na.omit(data[all.vars(formula)])
  if (all(is.na(response))) response <- eval(formula[[2]], data)
  is_categorical_response = class(response) == "factor" || class(response) == "logical"
  
  n <- nrow(data)
  scrambled_idxs <- sample(1:n)
  scrambled <- data[scrambled_idxs,]
  scrambled_response <- response[scrambled_idxs]
  cuts <- floor(seq(1, n, by = n / folds))
  cuts <- c(cuts, n)
  all_error <- c()
  
  for (i in 1:(length(cuts)-1)) {
    test_idxs <- cuts[i]:(cuts[i+1]-1)
    train <- scrambled[-test_idxs,]
    test <- scrambled[test_idxs,]
    mod <- fit_fun(formula, train)
    predicted <- predict_fun(mod, test)
    if (is_categorical_response) error <- predicted != scrambled_response[test_idxs]
    else error <- predicted - scrambled_response[test_idxs]
    all_error <- c(all_error, error)
  }
  mean(all_error^2)
}
```


## Overview
These are the formulae that we're using for later model fitting and analyses. We considered the number of days from the end of the poll to election day, the duration of the poll in days, the grade of the poll, whether the poll occurred on a national or state level, and which candidate the poll predicted in our analyses. In addition to these explanatory variables, we chose to investigate the differences between a binary response (whether the poll correctly predicted the outcome for its locale) and a continuous response (the 1-norm of the error between the poll's predicted percentages for each candidate and the actual percentage achieved by each candidate).

## Binary Response
```{r}
form1 <- is_correct ~ log(end_date_to_election) + log(samplesize) + is_national*predicted
summary(glm(form1, d, family = "binomial"))
```


```{r}
run_cv(d, form1, 
       fit_fun = function(formula, data) glm(formula, data, family = "binomial"),
       predict_fun = function(mod, data) predict(mod, data, type = "response") > 0.5)
run_cv(d, form1, 
       fit_fun = lda, 
       predict_fun = function(mod, data) predict(mod, data)$class)
```


```{r}
rf_data <- d %>% mutate(predicted = predicted == "Clinton")
rf_form <- as.factor(is_correct) ~ 
             end_date_to_election + samplesize + is_national + predicted
run_cv(rf_data, rf_form, fit_fun = randomForest)
```


```{r}
results <- sapply(1:20, function(k) {
  run_cv(d %>% mutate(predicted = ifelse(predicted == "Clinton", 1, 0)), form1, 
         fit_fun = function(formula, data) list(formula=formula, train=data),
         predict_fun = function(mod, data) {
           kNN(mod$formula, mod$train, data)
         })
})
plot(results, pch=20, ylim = c(0, 0.3))
results[which.min(results)]
```

## Continuous Response

```{r}
form2 <- error ~ log(end_date_to_election) + log(samplesize) + is_national*predicted
```


```{r}
run_cv(d, form2)
```

```{r}
rf_data2 <- d %>% mutate(predicted = predicted == "Clinton")
rf_form2 <- error ~ end_date_to_election + samplesize + is_national + predicted
run_cv(rf_data, rf_form, fit_fun = randomForest)

rf_fit <- randomForest(rf_form2, rf_data, importance = T, ntrees = 2000)
varImpPlot(rf_fit)
```
```{r}
plot(ctree(rf_form2, rf_data, 
           controls=cforest_control(mtry = 2, mincriterion = 0, maxdepth = 3)))
```

